Reparametrization of COM-Poisson Regression Models with Applications in the Analysis of Experimental Count Data \vspace{-0.4cm}

### 2018-08-27 presentations/ibc2018/slides-ibc2018.Rnw
Add presentations

\title[Reparametrization of COM-Poisson Regression Models]{
  Reparametrization of COM-Poisson Regression Models for Analysis of
  Count Data}

\author[Eduardo E. Ribeiro Jr]{
  Eduardo Elias Ribeiro Junior \inst{1} \inst{2} \\
  Walmes Marques Zeviani \inst{1} \\
  Wagner Hugo Bonat \inst{1} \\
  Clarice Garcia Borges Demétrio \inst{2} \\
  John Hinde \inst{3}
}

@@

\date{\small \today \\[0.2cm]
  \url{jreduardo@usp.br} | \url{edujrrib@gmail.com}}

@@

\begin{frame}[plain, noframenumbering]
  \titlepage
\end{frame}

@@

\begin{frame}{Poisson model and limitations}
  \vspace{-0.3cm}
  \begin{itemize}
    \item[] {\bf Count data}
    \item Number of times an event occurs in the observation unit;
    \item Mean-variance relationship.
      \vspace{0.3cm}
    \item[] {\bf GLM framework \citep{Nelder1972}}
    \item Provide suitable distribution for a counting random variables;
    \item Efficient algorithm for estimation and inference;
    \item Implemented in many software.
      \vspace{0.3cm}
    \item[] {\bf Poisson model}
    \item Relationship between mean and variance,
      $\text{E}(Y)=\text{Var}(Y)$;
      \vspace{0.3cm}
    \item[] {\bf Main limitations}
    \item Overdispersion (more common),
      $\text{E}(Y)<\text{Var}(Y)$
    \item Underdispersion (less common),
      $\text{E}(Y)>\text{Var}(Y)$
  \end{itemize}

@@

\begin{frame}{Weighted Poisson models}

  \begin{itemize}
  \item The family of weighted Poisson distributions (WPD)
    \citep{DelCastillo1998}, weights the Poisson probability function by
    a suitable function.
    \vspace{0.3cm}
  \item The probability mass function of the WPD is
    $$
    \Pr(Y = y) = \frac{\exp(-\lambda)\lambda^y}{y!}
    \frac{w(y)}{\text{E}_\lambda[w(Y)]}, \quad y \in \mathbb{N},
    $$
    where $\text{E}_\lambda(\cdot)$ denotes the mean value with respect
    to the Poisson random variable with parameter $\lambda$ and $w(y)$
    is a weight function.
    \vspace{0.3cm}
  \item The weight function may depend on extra parameter to ensure more
    flexibility to the distribution.
  \end{itemize}
\end{frame}

@@

  \begin{itemize}
  \item The COM-Poisson \citep{Shmueli2005} belongs to the family of
    weighted Poisson distributions with the weight function
    $w(y,\nu) = (y!)^{1-\nu}$.
    \vspace{0.3cm}
  \item   The probability mass function of $Y$ a
    COM-Poisson random variable is
    \begin{equation*}
      \label{eqn:pmf-compoisson}
      \Pr(Y = y) = \frac{\lambda^y \exp(-\lambda)}{
        (y!)^\nu \text{E}_\lambda[(Y!)^{1-\nu}]} =
      \frac{\lambda^y}{(y!)^\nu \text{Z}(\lambda, \nu)}, \quad
      \text{Z}(\lambda,\nu) = \sum_{j=0}^\infty \frac{\lambda^j}{(j!)^\nu},
    \end{equation*}
    where $\nu$ is the dispersion parameter.
    \vspace{0.3cm}
  \item The moments of distribution are not obtained in closed forms.
    There is approximations proposed by \citet{Shmueli2005,
      Sellers2010}:
    $$
    \text{E}(Y) \approx \dot{\text{E}}(Y) = \lambda^{1/\nu} - \frac{\nu
      - 1}{2\nu} \qquad \textrm{and} \qquad \text{Var}(Y) \approx
    \frac{\lambda^{1/\nu}}{\nu}.
    $$
  \end{itemize}
\end{frame}

@@

\begin{itemize}
  \item[] {\bf Model definition}
  \item Modelling the relationship between
    $\text{E}(Y_i)$ and $\bm{x}_i$ indirectly
    \citep{Sellers2010}; \\[-0.5cm]
    \begin{align*}
      Y_i \mid & \,\bm{x}_i \sim
                 \textrm{COM-Poisson}(\lambda_i,\, \nu) \\
               &\eta(\text{E}(Y_i \mid \bm{x}_i)) =
                 \log(\lambda_i) = \bm{x}_i^\top\bm{\beta} &
    \end{align*}
\end{itemize}
\vspace{0.3cm}

@@

\begin{itemize}
  \item[] {\bf Reparametrization}
  \item Introduced new parameter $\mu$, using the mean
    approximation\\[-0.2cm]
    \begin{equation*}
      \label{eqn:repar-cmp}
      \mu = \dot{\text{E}}(Y) = \lambda^{1/\nu} - \frac{\nu - 1}{2\nu}
      \quad \Rightarrow \quad
      \lambda = \left (\mu +
        \frac{(\nu - 1)}{2\nu} \right )^\nu;
    \end{equation*}
  \item Precision parameter is taken on the log scale to avoid
    restrictions on the parameter space\\[-0.3cm]
    $$\phi = \log(\nu) \Rightarrow \phi \in \mathbb{R};$$\\[0.3cm]
  \item[] {\bf Probability mass function}
  \item Replacing $\lambda$ and $\nu$ as function of $\mu$ and $\phi$ in
    the pmf of COM-Poisson\\[-0.2cm]
    \begin{equation*}
      \label{eqn:pmf-cmpmu}
      \Pr(Y=y \mid \mu, \phi) =
      \left ( \mu +\frac{ e^\phi-1}{2e^\phi} \right )^{ye^\phi}
      \frac{(y!)^{-e^\phi}}{Z(\mu, \phi)}.
    \end{equation*}
\end{itemize}

@@

\item[] {\bf Model definition}
\item Modelling relationship between $\dot{\text{E}}(Y_i)$ and
  $\bm{x}_i$
  directly\\[-0.5cm]
  \begin{align*}
    Y_i \mid & \,\bm{x}_i \sim
               \textrm{COM-Poisson}_\mu(\mu_i,\, \phi) \\
             &\log(\dot{\text{E}}(Y_i \mid \bm{x}_i)) =
               \log(\mu_i) = \bm{x}_i^\top\bm{\beta} &
  \end{align*}
  \vspace{-0.2cm}
\item[] {\bf Log-likelihood function
    ($\ell = \ell(\bm{\beta},\, \phi \mid \bm{y})$)}
  \vspace{0.05cm}
\item
  $
    \begin{aligned}[t]
    &\ell = e^\phi \left [
      \sum_{i=1}^n y_i
      \log \left( \mu_i + \frac{e^\phi-1}{2e^\phi} \right )
      -\sum_{i=1}^n \log(y_i!) \right ]
      -\sum_{i=1}^n \log(Z(\mu_i, \phi))\\
      &\text{\small where }\:
      \mu_i = \exp(\bm{x}_i^\top\bm{\beta})
      \end{aligned}
  $
\end{itemize}

@@

  The estimation and inference is based on the method of maximum
  likelihood. Let $\bm{\theta} = (\bm{\beta}, \phi)$ the model
  parameters. \vspace{0.2cm}
  \begin{itemize}
  \item Parameter estimates are obtained by numerical maximization of
    the log-likelihood function (by BFGS algorithm);\\[0.1cm]
    $\ell(\hat{\bm{\theta}}) =
    \text{max}\,\ell(\bm{\theta}),\:
    \bm{\theta} \in \mathbb{R}^{p+1}$;
    \vspace{0.2cm}
  \item Standard errors for regression coefficients are obtained based
    on the observed information matrix;\\[0.1cm]
    $\text{Var}(\hat{\bm{\theta}}) = -\mathcal{H}^{-1}$, where
    $\mathcal{H}$ is the matrix of second partial derivatives at
    $\hat{\bm{\theta}}$; \vspace{0.2cm}
  \item Confidence intervals for $\hat{\mu}_i$ are obtained by delta
    method.\\[0.1cm]
    $\text{Var}[g(\hat{\bm{\theta}})] \doteq \bm{G}\,
    \text{Var}(\hat{\bm{\theta}}) \bm{G}^\top $,
    where
    $\bm{G}^\top = (\partial g / \partial \beta_1,
    \ldots, \partial g / \partial \beta_p)^\top$;
    \vspace{0.2cm}
  \item The Hessian matrix $\mathcal{H}$ is obtained numerically by
    finite differences.
  \end{itemize}

@@

\begin{frame}{Orthogonality property of the MLEs}
\vspace{-0.3cm}

<<ortho-surf, fig.height=5, fig.width=9, fig.cap="Deviance surfaces contour plots under original and proposed parametrization. The ellipses are confidence regions (90, 95 and 99\\%), dotted lines are the maximum likelihood estimates, and points are the real parameters used in the simulation.">>=
include_graphics("figure/ortho-surf-1")
@

@@

  \begin{itemize}
  \item Three illustrative examples of count data analysis are
    reported.
    \begin{itemize} \normalsize
    \item Assessing toxicity of nitrofen in aquatic systems, an
      equidispersed example;
    \item Soil moisture and potassium doses on soybean culture, an
      overdispersed example; and
    \item Artificial defoliation in cotton phenology, an
      underdispersed example.
    \end{itemize}
    \vspace{0.3cm}
  \item In the data analysis, we consider:
    \begin{itemize}
    \item The COM-Poisson (original parametrization) model;
    \item The COM-Poisson (new parametrization) model;
    \item Quasi-Poisson model ($\text{Var}(Y) = \sigma
      \text{E}(Y)$); and
    \item The standard Poisson regression model.
    \end{itemize}
  \end{itemize}
\end{frame}

@@

\begin{frame}{Cotton bolls data}
\vspace{-0.2cm}
<<desc-cotton, fig.width=9, fig.height=5.8, fig.cap="">>=

knitr::include_graphics("figure/desc-cotton-1")

@@

  {\bf Alternative models:}
  % {\bf Modelling strategies:}
  \begin{itemize}
    \item Poisson ($\mu_{ij}$);
    \item COM-Poisson ($\lambda_{ij}=\eta(\mu_{ij})\,,\,\phi$)
    \item COM-Poisson$_\mu$ ($\mu_{ij}\,,\,\phi$)
    \item Quasi-Poisson (Var$(Y_{ij})=\sigma\mu_{ij}$)
  \end{itemize}

@@

\subsection{Soil moisture and potassium doses on soybean culture}

\begin{frame}{Soybean data}

  \begin{columns}
    \begin{column}{0.4\textwidth}
      \begin{figure}
        \includegraphics[width=4cm]{./images/soybean}
      \end{figure}
    \end{column}
    \begin{column}{0.6\textwidth}
      {\bf Aim:} evaluate the effects of potassium doses applied
      to soil in different soil moisture levels;\\[0.2cm]
      {\bf Design:} factorial 5 $\times$ 3 in a randomized complete
      block design (5 blocks);\\[0.2cm]
      {\bf Experimental unit:} a pot with a plant;\\[0.2cm]
      {\bf Factors:}
      \begin{itemize} \small
        \item Potassium fertilization dose (\texttt{K}):
        \item Soil moisture level (\texttt{umid}):
      \end{itemize}\vspace{0.1cm}
      {\bf Response variable:} Total number of bean seeds per pot;\\
    \end{column}
  \end{columns}

\begin{frame}{Soybean data}
\vspace{-0.2cm}
<<desc-soy, fig.width=9, fig.height=5.8, fig.cap="">>=

knitr::include_graphics("figure/desc-soy-1")

  {\bf Linear predictor:} {\small based on descriptive analysis,}
  \begin{itemize}
    \item $\log(\mu_{ijk}) = \beta_0 + \gamma_i + \tau_j +
      \beta_{1}\texttt{K}_k + \beta_{2}\texttt{K}_k^2 +
      \beta_{3j}\texttt{K}_k$\\[0.1cm]
    {\small
    $i$ varies according the blocks;\\
    $j$ varies in the levels of soil moisture;\\
    $k$ varies in the levels of potassium fertilization.}
  \end{itemize}

\begin{table}[!ht]
\centering \footnotesize
\caption{Parameter estimates (Est) and ratio between estimate and
  standard error (SE).}
\vspace{-0.1cm}
\label{tab:coef-soy}
\begin{tabular}{lrrrrrrrr}
  \toprule
  & \multicolumn{2}{c}{Poisson} &
    \multicolumn{2}{c}{COM-Poisson} &
    \multicolumn{2}{c}{COM-Poisson$_\mu$} &
    \multicolumn{2}{c}{Quasi-Poisson} \\
\cmidrule(lr){2-3} \cmidrule(lr){4-5} \cmidrule(lr){6-7} \cmidrule(lr){8-9}
% latex table generated in R 3.4.4 by xtable 1.8-2 package
% Wed Jun 27 14:44:34 2018
 & Est & Est/SE & Est & Est/SE & Est & Est/SE & Est & Est/SE \\
  \midrule
$\phi\,,\,\sigma$ &  &  & $-$0.779 & $-$4.721 & $-$0.782 & $-$4.737 & 2.615 &  \\
  $\beta_0$ & 4.867 & 144.289 & 2.232 & 6.042 & 4.867 & 97.781 & 4.867 & 89.225 \\
  $\gamma_{1}$ & $-$0.019 & $-$0.730 & $-$0.009 & $-$0.494 & $-$0.019 & $-$0.495 & $-$0.019 & $-$0.452 \\
  $\gamma_{2}$ & $-$0.037 & $-$1.373 & $-$0.017 & $-$0.921 & $-$0.037 & $-$0.931 & $-$0.037 & $-$0.849 \\
  $\gamma_{3}$ & $-$0.106 & $-$3.889 & $-$0.049 & $-$2.422 & $-$0.106 & $-$2.634 & $-$0.106 & $-$2.405 \\
  $\gamma_{4}$ & $-$0.092 & $-$3.300 & $-$0.042 & $-$2.102 & $-$0.092 & $-$2.237 & $-$0.092 & $-$2.040 \\
  $\tau_{1}$ & 0.132 & 3.647 & 0.061 & 2.295 & 0.132 & 2.472 & 0.132 & 2.255 \\
  $\tau_{2}$ & 0.124 & 3.432 & 0.057 & 2.177 & 0.124 & 2.326 & 0.124 & 2.122 \\
  $\beta_1$ & 0.616 & 11.014 & 0.284 & 4.729 & 0.616 & 7.464 & 0.616 & 6.811 \\
  $\beta_2$ & $-$0.276 & $-$10.250 & $-$0.127 & $-$4.589 & $-$0.276 & $-$6.946 & $-$0.276 & $-$6.338 \\
  $\beta_{31}$ & 0.146 & 4.268 & 0.067 & 2.614 & 0.146 & 2.892 & 0.146 & 2.639 \\
  $\beta_{32}$ & 0.165 & 4.829 & 0.076 & 2.884 & 0.165 & 3.272 & 0.165 & 2.986 \\
   \specialrule{0.01em}{0.3em}{0.3em}
 LogLik & \multicolumn{2}{c}{$-340.082$} & \multicolumn{2}{c}{$-325.241$} & \multicolumn{2}{c}{$-325.233$} & \multicolumn{2}{c}{$  -$}\\
 AIC & \multicolumn{2}{c}{$702.164$} & \multicolumn{2}{c}{$674.482$} & \multicolumn{2}{c}{$674.467$} & \multicolumn{2}{c}{$  -$}\\
 BIC & \multicolumn{2}{c}{$727.508$} & \multicolumn{2}{c}{$702.130$} & \multicolumn{2}{c}{$702.116$} & \multicolumn{2}{c}{$  -$} \\
  \bottomrule
\end{tabular}
\end{table}
\end{frame}

<<pred-soy, fig.height=3.8, fig.width=8, fig.cap="Dispersion diagrams of been seeds counts as function of potassium doses and humidity levels with fitted curves and confidence intervals (95\\%).">>=
include_graphics("figure/pred-soy-1")
@

\subsection{Assessing toxicity of nitrofen in aquatic systems}

\begin{frame}{Nitrofen data}

  \begin{columns}
    \begin{column}{0.4\textwidth}
      \begin{figure}
        \includegraphics[width=4cm]{./images/zooplankton}
      \end{figure}
    \end{column}
    \begin{column}{0.6\textwidth}
      {\bf Aim:} measure the reproductive toxicity of the herbicide
      nitrofen on a species of zooplankton
      (\textit{Ceriodaphnia dubia});\\[0.2cm]
      {\bf Design:} completely randomized design, with 10
      replicates;\\[0.2cm]
      {\bf Experimental unit:} zooplankton animal;\\[0.2cm]
      {\bf Factors:}
      \begin{itemize} \small
        \item herbicide nitrofen dose (\texttt{dose});
      \end{itemize}\vspace{0.1cm}
      {\bf Response variable:} Total number of live offspring;\\
    \end{column}
  \end{columns}

  \textbf{Linear predictors:}
  \begin{center}
    \begin{tabular}{ll}
      Linear:     &
          $\log(\mu_i) = \beta_0 + \beta_1 \texttt{dose}_i$,\\
      Quadratic: &
          $\log(\mu_i) = \beta_0 + \beta_1 \texttt{dose}_i +
             \beta_2 \texttt{dose}_i^2$ and\\
      Cubic:     &
          $\log(\mu_i) = \beta_0 + \beta_1 \texttt{dose}_i +
             \beta_2 \texttt{dose}_i^2 + \beta_3 \texttt{dose}_i^3.$
    \end{tabular}
  \end{center}

\begin{frame}{Likelihood ratio tests}

\vspace{-0.4cm}
\begin{table}[!ht]
\centering \footnotesize
\caption{Model fit measures and comparisons between linear predictors.}
\vspace{-0.2cm}
\label{tab:anova-ovos}
%\begin{tabularx}{\textwidth}{lCCCCCrC}
\begin{tabular}{lcccccrr}
\toprule
 Poisson & np & $\ell$ & AIC & 2(diff $\ell$) & diff np & P($>\rchi^2$) & \\
 \midrule
% latex table generated in R 3.4.4 by xtable 1.8-2 package
% Wed Jun 27 09:38:32 2018
 Linear & 2 & $-$180.667 & 365.335 &  &  &  &  \\
 Quadratic & 3 & $-$147.008 & 300.016 & 67.319 & 1 & 2.31E$-$16 &  \\
 Cubic & 4 & $-$144.090 & 296.180 & 5.835 & 1 & 1.57E$-$02 &  \\

\specialrule{0em}{0.5em}{0em} %% Apenas para espaçamento
  COM-Poisson & np & $\ell$ & AIC & 2(diff $\ell$) & diff np &
  P($>\rchi^2$) & $\hat{\phi}$ \\
  \midrule
% latex table generated in R 3.4.4 by xtable 1.8-2 package
% Wed Jun 27 09:38:32 2018
 Linear & 3 & $-$167.954 & 341.908 &  &  &  & $-$0.893 \\
  Quadratic & 4 & $-$146.964 & 301.929 & 41.980 & 1 & 9.22E$-$11 & $-$0.059 \\
  Cubic & 5 & $-$144.064 & 298.129 & 5.800 & 1 & 1.60E$-$02 & 0.048 \\

\specialrule{0em}{0.5em}{0em} %% Apenas para espaçamento
  COM-Poisson$_\mu$ & np & $\ell$ & AIC & 2(diff $\ell$) & diff np &
  P($>\rchi^2$) & $\hat{\phi}$ \\
  \midrule
% latex table generated in R 3.4.4 by xtable 1.8-2 package
% Wed Jun 27 09:38:32 2018
 Linear & 3 & $-$167.652 & 341.305 &  &  &  & $-$0.905 \\
  Quadratic & 4 & $-$146.950 & 301.900 & 41.405 & 1 & 1.24E$-$10 & $-$0.069 \\
  Cubic & 5 & $-$144.064 & 298.127 & 5.773 & 1 & 1.63E$-$02 & 0.047 \\

\specialrule{0em}{0.5em}{0em} %% Apenas para espaçamento
  Quasi-Poisson & np & QDev & AIC & F & diff np & P($>F$) & $\hat{\sigma}$ \\
  \midrule
% latex table generated in R 3.4.4 by xtable 1.8-2 package
% Wed Jun 27 09:38:32 2018
 Linear  & 3 & 123.929 &  &  &  &  & 2.262 \\
  Quadratic & 4 & 56.610 &  & 60.840 & 1 & 5.07E$-$10 & 1.106 \\
  Cubic & 5 & 50.774 &  & 5.659 & 1 & 2.16E$-$02 & 1.031 \\

 \bottomrule
%\end{tabularx}
\end{tabular}
\end{table}

\begin{frame}{Parameter estimates and fitted values}

  \vspace{-0.4cm}
\begin{table}[!ht]
\centering \footnotesize
\caption{Parameter estimates (Est) and ratio between estimate and
  standard error (SE).}
\vspace{-0.1cm}
\label{tab:coef-ovos}
\begin{tabular}{lrrrrrrrr}
  \toprule
  & \multicolumn{2}{c}{Poisson} &
    \multicolumn{2}{c}{COM-Poisson} &
    \multicolumn{2}{c}{COM-Poisson$_\mu$} &
    \multicolumn{2}{c}{Quasi-Poisson} \\
\cmidrule(lr){2-3} \cmidrule(lr){4-5} \cmidrule(lr){6-7} \cmidrule(lr){8-9}
% latex table generated in R 3.4.4 by xtable 1.8-2 package
% Wed Jun 27 09:38:32 2018
 & Est & Est/SE & Est & Est/SE & Est & Est/SE & Est & Est/SE \\
  \midrule
$\beta_{0}$ & 3.477 & 62.817 & 3.649 & 4.850 & 3.477 & 64.308 & 3.477 & 61.860 \\
  $\beta_{1}$ & $-$0.086 & $-$0.433 & $-$0.091 & $-$0.448 & $-$0.088 & $-$0.452 & $-$0.086 & $-$0.426 \\
  $\beta_{2}$ & 0.153 & 0.863 & 0.161 & 0.878 & 0.155 & 0.894 & 0.153 & 0.850 \\
  $\beta_{3}$ & $-$0.097 & $-$2.398 & $-$0.102 & $-$2.229 & $-$0.098 & $-$2.464 & $-$0.097 & $-$2.361 \\
   \bottomrule
\end{tabular}
\end{table}
  \vspace{-0.4cm}

<<pred-ovos, fig.height=3.5, fig.width=5.5, out.width="0.8\\textwidth", fig.cap="Number of live offsprings observed for each nitrofen concentration level with fitted curves and 95\\% confidence intervals.">>=
include_graphics("figure/pred-ovos-1")
@

\subsection*{}

\begin{frame}{Additional results}

@@

<<comp-times, fig.height=4, fig.width=8, fig.cap="Computational times to fit the models under original and reparametrized versions based on the fifty repetitions.">>=

library(latticeExtra)
source("../../codes/lattice-panels.R")

benchs <- readRDS("../../codes/comparetimes.rds")
xlabs <- expression("COM-Poisson", "COM-Poisson"[mu])
bwplot(time ~ expr | case,
       ylab = "Time (seconds)",
       scales = list(
           y = list(relation = "free"),
           x = list(at = 1:2, labels = xlabs)
       ),
       data = benchs)

>>>>

\title[Reparametrization of COM-Poisson Regression Models]{
  Reparametrization of COM-Poisson Regression Models with Applications
  in the Analysis of Experimental Count Data \vspace{-0.4cm}}

\author[Clarice Garcia Borges Demétrio]{
  Eduardo Elias Ribeiro Junior \inst{1} \inst{2} \\
  Walmes Marques Zeviani \inst{1} \\
  Wagner Hugo Bonat \inst{1} \\
  Clarice Garcia Borges Demétrio \inst{2} \\
  John Hinde \inst{3}
}

@@

\date{\small 10th July 2018 \\
  \url{jreduardo@usp.br} | \url{edujrrib@gmail.com}}

@@

\begin{frame}[plain, noframenumbering]
  \begin{center}
    \bfseries
    XXIX\textsuperscript{TH} INTERNATIONAL BIOMETRIC CONFERENCE\\
    Barcelona, Spain, 8-13 July 2018
  \end{center}
  \vspace{-0.5cm}
  \titlepage
\end{frame}

@@

\begin{frame}{Count data}

  Number of times an event occurs in the observation unit.\\[0.3cm]

  Random variables that assume non-negative integer values.\\[0.3cm]

  Let $Y$ be a counting random variable, so that $y=0,1,2,\ldots$\\[0.5cm]

  Examples in experimental researches:

  \begin{itemize}
    \item number of grains produced by a plant;
    \item number of fruits produced by a tree;
    \item number of insects on a particular cell;
    \item others.
  \end{itemize}

@@

\begin{frame}{Poisson model and limitations}

  \begin{itemize}
    \item[] {\bf GLM framework \citep{Nelder1972}}
    \item Provide suitable distribution for a counting random variables;
    \item Efficient algorithm for estimation and inference;
    \item Implemented in many software.
      \vspace{0.5cm}
    \item[] {\bf Poisson model}
    \item Relationship between mean and variance,
      $\text{E}(Y)=\text{Var}(Y)$;
      \vspace{0.5cm}
    \item[] {\bf Main limitations}
    \item Overdispersion (more common),
      $\text{E}(Y)<\text{Var}(Y)$
    \item Underdispersion (less common),
      $\text{E}(Y)>\text{Var}(Y)$
  \end{itemize}

@@

  \begin{itemize}
    \itemsep7.5pt\parskip0pt\parsep0pt
  \item Probability mass function \citep{Shmueli2005} takes the form \\
    \begin{equation*} \small
    \label{eqn:pmf-cmp}
    \Pr(Y = y \mid \lambda, \nu) =
      \frac{\lambda^y}{(y!)^\nu Z(\lambda, \nu)}\,,
    \qquad Z(\lambda, \nu) = \sum_{j=0}^\infty
      \frac{\lambda^j}{(j!)^\nu}\,,
  \end{equation*}
    where $\lambda > 0$ and $\nu \geq 0$.
  \item Moments are not available in closed form;
  \item Expectation and variance can be closely approximated by
    \begin{equation*}
      \label{eqn:mean-aprox}
      \text{E}(Y) \approx \lambda^{1/\nu} - \frac{\nu - 1}{2\nu} \qquad
      \textrm{and} \qquad
      \text{Var}(Y) \approx \frac{\lambda^{1/\nu}}{\nu}
    \end{equation*}
    with accurate approximations for $\nu \leq 1$ or $\lambda > 10^\nu$
    \citep{Shmueli2005, Sellers2012}.
  \end{itemize}

@@

\begin{itemize}
  \item[] {\bf Model definition}
  \item Modelling the relationship between
    $\text{E}(Y_i)$ and $\bm{x}_i$ indirectly
    \citep{Sellers2010}; \\[-0.5cm]
    \begin{align*}
      Y_i \mid & \,\bm{x}_i \sim
                 \textrm{COM-Poisson}(\lambda_i,\, \nu) \\
               &\eta(E(Y_i \mid \bm{x}_i)) =
                 \log(\lambda_i) = \bm{x}_i^\top\bm{\beta} &
    \end{align*}
\end{itemize}
\vspace{0.3cm}

@@

\begin{itemize}
  \item[] {\bf Reparametrization}
  \item Introduced new parameter $\mu$, using the mean
    approximation\\[-0.2cm]
    \begin{equation*}
      \label{eqn:repar-cmp}
      \mu = \lambda^{1/\nu} - \frac{\nu - 1}{2\nu}
      \quad \Rightarrow \quad
      \lambda = \left (\mu +
        \frac{(\nu - 1)}{2\nu} \right )^\nu;
    \end{equation*}
  \item Precision parameter is taken on the log scale to avoid
    restrictions on the parameter space\\[-0.3cm]
    $$\phi = \log(\nu) \Rightarrow \phi \in \mathbb{R};$$\\[0.3cm]
  \item[] {\bf Probability mass function}
  \item Replacing $\lambda$ and $\nu$ as function of $\mu$ and $\phi$ in
    the pmf of COM-Poisson\\[-0.2cm]
    \begin{equation*}
      \label{eqn:pmf-cmpmu}
      \Pr(Y=y \mid \mu, \phi) =
      \left ( \mu +\frac{ e^\phi-1}{2e^\phi} \right )^{ye^\phi}
      \frac{(y!)^{-e^\phi}}{Z(\mu, \phi)}.
    \end{equation*}
\end{itemize}

@@

\item[] {\bf Model definition}
\item Modelling relationship between $E(Y_i)$ and $\bm{x}_i$
  directly\\[-0.5cm]
  \begin{align*}
    Y_i \mid & \,\bm{x}_i \sim
               \textrm{COM-Poisson}_\mu(\mu_i,\, \phi) \\
             &\log(E(Y_i \mid \bm{x}_i)) =
               \log(\mu_i) = \bm{x}_i^\top\bm{\beta} &
  \end{align*}
  \vspace{-0.2cm}
\item[] {\bf Log-likelihood function
    ($\ell = \ell(\bm{\beta},\, \phi \mid \bm{y})$)}
  \vspace{0.05cm}
\item
  $
    \begin{aligned}[t]
    &\ell = e^\phi \left [
      \sum_{i=1}^n y_i
      \log \left( \mu_i + \frac{e^\phi-1}{2e^\phi} \right )
      -\sum_{i=1}^n \log(y_i!) \right ]
      -\sum_{i=1}^n \log(Z(\mu_i, \phi))\\
      &\text{\small where }\:
      \mu_i = \exp(\bm{x}_i^\top\bm{\beta})
      \end{aligned}
  $
\end{itemize}

@@

  The estimation and inference is based on the method of maximum
  likelihood. Let $\bm{\theta} = (\bm{\beta}, \phi)$ the model
  parameters. \vspace{0.2cm}
  \begin{itemize}
  \item Parameter estimates are obtained by numerical maximization of
    the log-likelihood function (by BFGS algorithm);\\[0.1cm]
    $\ell(\hat{\bm{\theta}}) =
    \text{max}\,\ell(\bm{\theta}),\:
    \bm{\theta} \in \mathbb{R}^{p+1}$;
    \vspace{0.2cm}
  \item Standard errors for regression coefficients are obtained based
    on the observed information matrix;\\[0.1cm]
    $\text{Var}(\hat{\bm{\theta}}) = -\mathcal{H}^{-1}$, where
    $\mathcal{H}$ is the matrix of second partial derivatives at
    $\hat{\bm{\theta}}$; \vspace{0.2cm}
  \item Confidence intervals for $\hat{\mu}_i$ are obtained by delta
    method.\\[0.1cm]
    $\text{Var}[g(\hat{\bm{\theta}})] \doteq \bm{G}\,
    \text{Var}(\hat{\bm{\theta}}) \bm{G}^\top $,
    where
    $\bm{G}^\top = (\partial g / \partial \beta_1,
    \ldots, \partial g / \partial \beta_p)^\top$;
  \item The Hessian matrix $\mathcal{H}$ is obtained numerically by
    finite differences.
  \end{itemize}

@@

\begin{frame}{Orthogonality property of the MLEs}
\vspace{-0.3cm}
<<ortho-surf, fig.height=5, fig.width=9, fig.cap="Deviance surfaces contour plots under original and proposed parametrization. The ellipses are confidence regions (90, 95 and 99\\%), dotted lines are the maximum likelihood estimates, and points are the real parameters used in the simulation.">>=
include_graphics("figure/ortho-surf-1")
@

@@

  \begin{itemize}
    \item Three illustrative examples of count data analysis are
      reported.
      \begin{itemize} \normalsize
      \item Assessing toxicity of nitrofen in aquatic systems, an
        equidispersed example;
      \item Soil moisture and potassium doses on soybean culture, an
        overdispersed example; and
      \item Artificial defoliation in cotton phenology, an
        underdispersed example.
      \end{itemize}
    \item In the data analysis, we consider the COM-Poisson model in the
      two forms (original and new parametrization) and the quasi-Poisson
      regression model as alternative models for the standard Poisson
      regression model.
  \end{itemize}
\end{frame}

@@

@@

@@

\subsection*{Additional results}

\begin{frame}{Computational times and orthogonality property}

@@

<<comp-times, fig.height=4, fig.width=8, fig.cap="Computational times to fit the models under original and reparametrized versions based on the fifty repetitions.">>=
include_graphics("figure/comp-times-1")
@
